---
title: "Dirichlet distribution tools"
author: "Lorenzo Gaborini"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dirichlet distribution tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
\newcommand{\simplex}[1]{\text{Simplex}{(#1)}}
\newcommand{\dirichlet}[1]{\text{Dirichlet}{(#1)}}
\renewcommand{\vec}[1]{\boldsymbol{#1}}
\newcommand{\EE}{\mathop{\mathbb{E}}}
\newcommand{\Var}{\mathop{\mathrm{Var}}}

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(rsamplestudy)
library(magrittr)
```

This package contains functions to sample from a Dirichlet distribution.


## Notation


We call $\simplex{p-1}$ the set of vectors in $\mathbb{R}^p$ whose components form a distribution on $p$ items.   
In other words, $\vec{X} \in \simplex{p-1}$ iff $X_i \in [0,1]$ and $\sum_{i=1}^p X_i = 1$.

Let $\vec{X}$ be a vector in $\simplex{p-1}$, let $\vec{\alpha}$ be a vector in $\mathbb{R}^p$ with positive components.

Then, $X$ can have the Dirichlet distribution: $X \mid \vec{\alpha} \sim \dirichlet{\vec{\alpha}}$.

We can write $\vec{\alpha} = \alpha_0 \, \vec{\nu}$, where:

* $\alpha_0 = \sum_i \alpha_i$ is the **concentration parameter**: $\alpha_0 \in \mathbb{R}^{+}$
* $\vec{\nu}$ is the **base measure**: $\vec{\nu} \in \simplex{p-1}$.

  
## Properties

- $\EE[\vec{X}] = \vec{\alpha} / \alpha_0 = \vec{\nu}$
- $\Var[\vec{X}] = \dfrac{ \vec{\nu} \cdot (1 - \vec{\nu})}{\alpha_0 + 1}$


## Particular cases

If $\vec{\nu}$ is constant, $\vec{\nu} = 1/p$. The Dirichlet distribution is called **symmetric Dirichlet**, and is only parametrised by the concentration parameter $\alpha_0$.

- When $\alpha_0 = p$, $\alpha_i = 1 \forall i$: the Dirichlet distribution is **uniform** over $\simplex{p-1}$.  This is the least informative situation.
- When $\alpha_0 \to 0$, $\alpha_i \to 0$: the Dirichlet distribution concentrates over the border of $\simplex{p-1}$.
- When $\alpha_0 \gg 1$, $\alpha_i \to 0$: the Dirichlet distribution concentrates over the mean (and the variance decreases).

## Examples

Sample from a $p=4$ Dirichlet:

```{r}
n <- 100

# The parameter
alpha <- c(0.5, 0.1, 5, 0.1)
p <- length(alpha)

# Concentration and base measure
alpha0 <- sum(alpha)
nu <- alpha / alpha0
```

$\alpha_0 = `r alpha0`$   
$\vec{\nu} = \left( `r nu` \right)$   
$\vec{\alpha} = \left( `r alpha` \right)$

```{r}
df_diri <- fun_rdirichlet(n, alpha)
head(df_diri)
```

```{r sym-sens-plot-comp, echo = FALSE}

source('scatter_matrix_simplex.R')
scatter_matrix_simplex(df_diri)

```

### Uniform sampling

Uniform base measure, unitary Dirichlet parameters: $\vec{\alpha} = \vec{1}$.   

```{r}
nu <- rep(1, p)/p
alpha0 <- p
alpha <- nu * alpha0
```

$\alpha_0 = `r alpha0`$   
$\vec{\nu} = \left( `r nu` \right)$   
$\vec{\alpha} = \left( `r alpha` \right)$

```{r}
df_diri <- fun_rdirichlet(n, alpha)
head(df_diri)
```

```{r}
scatter_matrix_simplex(df_diri)
```


### Concentrated sampling

Concentration parameter $\gg p$, Dirichlet parameters $\gg 1$

```{r}
nu <- rep(1, p)/p
alpha0 <- 100
alpha <- nu * alpha0
```

$\alpha_0 = `r alpha0`$   
$\vec{\nu} = \left( `r nu` \right)$   
$\vec{\alpha} = \left( `r alpha` \right)$

```{r}
df_diri <- fun_rdirichlet(n, alpha)
head(df_diri)
```

```{r}
scatter_matrix_simplex(df_diri)
```


### Degenerate sampling

Concentration parameter $\\ p$, Dirichlet parameters $\\ 1$

```{r}
nu <- rep(1, p)/p
alpha0 <- 0.01
alpha <- nu * alpha0
```

$\alpha_0 = `r alpha0`$   
$\vec{\nu} = \left( `r nu` \right)$   
$\vec{\alpha} = \left( `r alpha` \right)$

```{r}
df_diri <- fun_rdirichlet(n, alpha)
head(df_diri)
```

```{r}
scatter_matrix_simplex(df_diri)
```
