---
title: "Dirichlet-Dirichlet study generation"
author: "Lorenzo Gaborini"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dirichlet-Dirichlet study generation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
\newcommand{\simplex}[1]{\text{Simplex}{(#1)}}
\newcommand{\dirichlet}[1]{\text{Dirichlet}{(#1)}}
\renewcommand{\vec}[1]{\boldsymbol{#1}}
\newcommand{\EE}{\mathop{\mathbb{E}}}
\newcommand{\Var}{\mathop{\mathrm{Var}}}

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(purrr)
library(rsamplestudy)
source('scatter_matrix_simplex.R')
```

A **study** is the set of parameters in a model, data generated from those parameters (**population**), and three sets of samples: the reference/questioned/background items.

This package implements the generation of selected studies.   
This vignette describes the Dirichlet-Dirichlet model.

## The model

(To be described).




## Population generation

The population can be generated using `fun_rdirichlet_population`:

```{r}
# Population parameters:
# Number of sources
n <- 10
# Number of items per source
m <- 20
# Number of observations per item
p <- 4

list_pop <- fun_rdirichlet_population(n, m, p)
```

The output contains:

- the population `df_pop`
- the model parameters `df_sources` and `alpha`
- names of the components of the source vectors `names_source`
- names of the components of the item vectors `names_var`

Notice that the hyperparameter is sampled, too (but it can be fixed).

```{r}
head(list_pop$df_pop)
head(list_pop$df_sources)
```

## Hyperparameters

We assume that the Dirichlet hyperparameter (the level farther from the data) comes from the Uniform distribution on the (p-1)-Simplex.   
In other words, we will sample the Dirichlet hyperparameter from the $\dirichlet{\vec{1}}$ distribution. 

The shortcut function the package is `fun_rdirichlet_hyperparameter`:

```{r}
df_diri <- map_dfr(1:300, ~ fun_rdirichlet_hyperparameter(3))
scatter_matrix_simplex(df_diri)
```


## Partitioning

Once the population is generated, the reference/questioned/background samples must be extracted.   
This is generically done using `make_dataset_splits`:

```{r}

k_ref <- 10
k_quest <- 5

list_samples <- make_dataset_splits(list_pop$df_pop, k_ref, k_quest)
names(list_samples)
```

```{r}
head(list_samples$df_reference)
head(list_samples$df_questioned)
head(list_samples$df_background)
```

### Source parameters

The chosen sources can be fixed. 

See the documentation for `make_dataset_splits`.




