---
title: "Sampling: reference and questioned items"
author: "Lorenzo Gaborini"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sampling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette introduces the concepts behind the comparison of two sets of items, as the functions `make_dataset_splits` and `make_idx_splits` implement.

## Source data

Suppose that we have a set of observations with their associated source.    
The $j$-th observation from the $i$-th source is indicated with $X_{ij}$.
In total, we have $m$ sources, $n_i$ observations per source.

The general model for the data is:

$$X_{ij} \sim f(\theta_i) \quad \text{iid } j = 1, ..., n_i $$

For the purpose of illustration, let's use the popular dataset `chickwts`:

```{r}
data(chickwts)
head(chickwts)
```

The data are the recorded weights of `r nrow(chickwts)` chicks according to `nlevels(chickwts$feed)` different feed supplements (`feed` column).
We consider `feed` as the source label: we assume that chicks are exchangeable given the administered feed.

## Two-sample concept

This package is used in a forensic evaluative setting, where one has a set of observations from a known source (the **reference set** and the **reference source**), and a set of observations whose source(s) is (are) unknown (the **questioned set**).    

One commonly states two competing hypotheses, e.g. whether the source of the questioned set is the reference source, or one (or more) different sources.   
These hypotheses are usually named $H_1$ and $H_2$. Notice that $H_2$ can consider a single alternative source, or multiple sources: all of them are unknown and different from the reference one. 

This package assists with the generation of the reference and questioned set starting from the observed data.frame.   

The dimensions of the reference and questioned sets can be specified, and are usually much smaller than the full data.    
It follows that the rows which have not been picked constitute a third set, the **background set**.    
This set is used in a Bayesian setting to learn the (hyper)priors for the statistical model.

## Sample generation

To connect with the example, we consider `horsebean` feed as the reference source. From the full data, we pick 5 observations to constitute each one of the reference and questioned sets. Let's see how `rsamplestudy` does it:

```{r}
set.seed(123)

library(rsamplestudy)

n_items <- 5
col_source <- 'feed'

list_split <- make_dataset_splits(chickwts, k_ref = n_items, k_quest = n_items, col_source = 'feed', source_ref = 'horsebean')
list_split
```

The return value contains three lists: the indexes of the rows in `chickwts`, and three data frames (i.e. `chickwts` split up across the previous indexes).

### Questioned source choice 

Notice that it automatically picked a questioned source.    
This behaviour is specified in `make_dataset_splits` documentation: if the questioned source is not specified, then any source different from the reference one is picked as a potential source. Then, items are randomly picked.

We can also force the questioned source(s): e.g. we pick items from `'horsebean'` and `'soybean'`

```{r}
list_split <- make_dataset_splits(chickwts, k_ref = n_items, k_quest = n_items, col_source = 'feed', 
                                  source_ref = 'horsebean', source_quest = c('horsebean', 'soybean'))
list_split
```

### Sampling with repetition

It is always guaranteed that no item appears more than once across sets.   
It may be explicitely allowed to pick multiple times an item **in** a set (but it will never appear in other sets).

For example, chick counts per feed:

```{r}
table(chickwts$feed)
```

We might want to obtain bootstrap samples from chicken who have been fed with horsebeans, by sampling 100 times.   


```{r}
n_items_rep <- 100
list_split <- make_dataset_splits(chickwts, 
                    k_ref = n_items_rep, 
                    k_quest = n_items_rep, col_source = 'feed', source_ref = 'horsebean')

head(list_split$df_reference)
head(list_split$df_questioned)
head(list_split$df_background)
```

Items are never picked more than once across the sets:

```{r}
intersect(list_split$idx_reference, list_split$idx_questioned)
intersect(list_split$idx_questioned, list_split$idx_background)
intersect(list_split$idx_reference, list_split$idx_background)
```


### Background selection

The background set can be constituted in two ways depending if the **potential sampled sources** (i.e. sources whose items *could* appear either in the reference or in the questioned set) are excluded or not.

`make_dataset_splits` accepts the parameter `background` which can take two values:

`'outside'` (default): the potential observed sources may appear in the background dataset.

```{r}
set.seed(123)
list_split <- make_dataset_splits(chickwts, k_ref = n_items, k_quest = n_items, col_source = 'feed', source_ref = 'horsebean',
                                  background = 'outside')

# Observed reference source
unique(list_split$df_reference$feed)
# Observed questioned sources
unique(list_split$df_questioned$feed)
# Sources in background
unique(list_split$df_background$feed)

```

`'others'`: the potential observed sources cannot appear in the background set.

Notice that if the potential sources span the entire available sources, than the background dataset must be empty.
It is the case when the questioned source is not specified, as it is automatically considered to be "all but the reference source".


```{r}
set.seed(123)
list_split <- make_dataset_splits(chickwts, k_ref = n_items, k_quest = n_items, col_source = 'feed', 
                                  source_ref = 'horsebean', source_quest = NULL,
                                  background = 'others')

# Observed reference source
unique(list_split$df_reference$feed)
# Observed questioned sources
unique(list_split$df_questioned$feed)
# Sources in background
unique(list_split$df_background$feed)

```

Once the potential observed sources are restricted, the background selection matters:

```{r}
set.seed(123)
list_split <- make_dataset_splits(chickwts, k_ref = n_items, k_quest = n_items, col_source = 'feed', 
                                  source_ref = 'horsebean', source_quest = c('horsebean', 'soybean'),
                                  background = 'others')

# Observed reference source
unique(list_split$df_reference$feed)
# Observed questioned sources
unique(list_split$df_questioned$feed)
# Sources in background
unique(list_split$df_background$feed)

```
